A Step-Indexed Logic in Agda

### 2023-09-22 papers/July2023/paper.tex
updates to section 3

The authors are interested in using step-indexed logical relations for
purposes of developing mechanized proofs in Agda of the Gradual
Guarantee~\citep{Siek:2015ac} and Parametricity~\citep{REYNOLDS74C}
for gradually-typed polymorphic
languages~\citep{Ahmed:2011fk,Ahmed:2017aa,Igarashi:2017aa,New:2019ab,Labrada:2020tk}.
One option is to use Guarded Cubical Agda~\citep{Veltri:2020aa}, an
experimental extension to Agda, and another option is to develop an
embedding of a step-indexed logic in (vanilla) Agda. The authors are
interested in both approaches, as there are tradeoffs that should be
evaluated. For example, using Cubical Agda comes with the down-side of
losing Uniqueness of Identity Proofs (UIP) but on the other hand
Guarded Cubical Agda provides a more ergonomic user-interface for
step-indexed reasoning than an embedded logic. The authors chose to
first investigate the embedding approach, which we describe in this
paper.

@@

@@

UNDER CONSTRUCTION

To evaluate the utility of SIL, Section~\ref{sec:STLC} presents a case
study of proving semantic type safety for the Simply Typed Lambda
Calculus extended with recursive functions. We find that SIL
streamlines the definition of logical relations and is helpful in many
of the proofs, but there is room for improvement regarding the
ergonomics of interactive proof development using SIL.  Limitations in
Agda's inference algorithm sometimes leads to verbosity in SIL proofs.

@@

\input{latex/July2023/STLC2.tex}
\input{latex/July2023/LogRel2.tex}
\input{latex/July2023/STLCDeterministic2.tex}
\input{latex/July2023/STLCBind2.tex}
\input{latex/July2023/STLCSafe2.tex}
\input{latex/July2023/STLCTypeSafe.tex}

@@

We organize the rest of the article as follows. We review Appel an
McAllester's work in Section~\ref{sec:appel-mcallester}, then
introduce our representation of step-indexed propositions and
predicates in Agda in Sections \ref{sec:propositions} and
\ref{sec:predicates}.  We formalize several important concepts in
Section~\ref{sec:fun-approx-iter}: functionals, approximation, and
iteration, and prove some basic results about them.  We construct all
of the logical connectives of SIL in
Section~\ref{sec:open-propositions}, including the definition of
recursive predicates and their fixpoint theorem.  The proof theory of
SIL is developed in Section~\ref{sec:proofs}, where we construct the
proof rules of this modal logic.

@@

>>>>

The authors are interested in using step-indexed logical relations for
purposes of developing mechanized proofs in Agda of the Gradual
Guarantee~\citep{Siek:2015ac} and Parametricity~\citep{REYNOLDS74C}
for gradually-typed polymorphic
languages~\citep{Ahmed:2011fk,Ahmed:2017aa,Igarashi:2017aa,New:2019ab,Labrada:2020tk}.
One option is to use Guarded Cubical Agda~\citep{Veltri:2020aa}, an
experimental extension to Agda, and another option is to develop an
embedding of a step-indexed logic in (vanilla) Agda. The authors are
interested in both approaches, as there are tradeoffs that should be
evaluated. For example, using Cubical Agda comes with the down-side of
losing Uniqueness of Identity Proofs (UIP) but on the other hand
Guarded Cubical Agda provides a more ergonomic user-interface for
step-indexed reasoning than what's possible in an embedded logic. The
authors chose to first investigate the embedding approach.

@@

We describe SIL from the user perspective in
§\ref{sec:SIL-intro}, including what kinnds of formulas can be
written in SIL and how to conduct proofs in SIL. As a step-indexed
logic, SIL supports user-defined recursive predicates and a ``later''
operator.

@@

To evaluate the utility of SIL, §\ref{sec:STLC} presents a case
study of proving semantic type safety for the Simply Typed Lambda
Calculus extended with recursive functions. We find that SIL
streamlines the definition of logical relations and is helpful in many
of the proofs, but there is room for improvement regarding the
ergonomics of interactive proof development using SIL.  Limitations in
Agda's inference algorithm sometimes leads to verbosity in SIL proofs.

TODO: roadmap for the implementation of SIL

@@

\input{latex/July2023/STLC2.tex}
\input{latex/July2023/LogRel2.tex}
\input{latex/July2023/STLCBind2.tex}
\input{latex/July2023/STLCSafe2.tex}
\input{latex/July2023/STLCTypeSafe.tex}

@@

We organize the rest of the article as follows. We review Appel an
McAllester's work in §\ref{sec:appel-mcallester}, then
introduce our representation of step-indexed propositions and
predicates in Agda in Sections \ref{sec:propositions} and
\ref{sec:predicates}.  We formalize several important concepts in
§\ref{sec:fun-approx-iter}: functionals, approximation, and
iteration, and prove some basic results about them.  We construct all
of the logical connectives of SIL in
§\ref{sec:open-propositions}, including the definition of
recursive predicates and their fixpoint theorem.  The proof theory of
SIL is developed in §\ref{sec:proofs}, where we construct the
proof rules of this modal logic.

@@

\input{latex/July2023/STLCDeterministic2.tex}
